 

### 2021-02-01 vertical-thermal-review.Rmd
organization

**Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance , and the implications for understanding forest ecosystem responses and feedbacks to climate change .**
We structure the review around six major questions (Fig. `r fig_schematic`).

@@

vi.	What are the implications for modeling/ scaling / What are the implications of these leaf-scale responses at the ecosystem scale [in the context of climate change] / How do these leaf-level responses scale to entire ecosystems(*nerge these better*)

@@

### Carbon dioxide
**Carbon dioxide (CO~2~) concentrations tend to be higher in the understory, primarily at night.** 
	Higher in the understory, particularly at dusk [@koike_leaf_2001]. Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999]. Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]
Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy. 
~~Furthermore, increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.~~ (*but make the point about big tree mort elsewhere*)

(*keep discussion of CO2 to what's in the paragraph above. Definitely don't want to go beyond a brief mention of warming effects on soil respiration*): (Elsa: see these papers on the long-term Harvard Forest soil warming experiments. CO2 emissions increased, then decreased, then increased again. It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J. and Frey, S.D., 2015. Long-term forest soil warming alters microbial communities in temperate forest soils. Frontiers in microbiology, 6, p.104.
- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T. and Morrisseau, S., 2002. Soil warming and carbon-cycle feedbacks to the climate system. Science, 298(5601), pp.2173-2176.
- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A. and Grandy, A.S., 2017. Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world. Science, 358(6359), pp.101-105.) 
- Nottingham, A.T., Meir, P., Velasquez, E. and Turner, B.L., 2020. Soil carbon loss by experimental warming in a tropical forest. Nature, 584(7820), pp.234-237.

@@

## Implications for Modeling and projecting climate change responses

@@

*(@Nidhi & @Krista, feel free to edit/remove any of the above text. I'm adding more text than we'll want in the end, but I want to air on the side of more rather than less so that you can make decisions about what content you think best fits the goal of this review. At some point, perhaps we should discuss this section further and I can revise based on that.) *

@@

### Implications for climate change responses

@@

>>>>

**Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance , and the implications for understanding forest ecosystem responses and feedbacks to climate change .**
We structure the review around five major questions (Fig. `r fig_schematic`).

@@

We then consider the implications for modeling/ scaling /climate change responses.

@@

### Carbon dioxide

**Atmopsheric carbon dioxide (CO~2~) concentrations tend to be higher near ground level at night, but largely dissipate during the day [@koike_leaf_2001; @yang_spatial_1999].** 
~~concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]~~ *(KAT: I'm confused by this. We want to discuss just atmospheric CO2 here, so I think this doesn't fit)*
Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
~~If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy.~~ *(this last sentence could be cut for space)*

@@

## Implications

*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*

@@

@@

### Climate change responses

@@

increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness
